---
title: "Covid Final"
output: github_document
date: "2026-06-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(usmap)
library(lubridate)
```

## Question(s) of interest
1. Where are cases located in the US?
2. How does the US compare to the rest of the world by Covid deaths?
3. How much of an impact is just population size in terms of US deaths?

## Read in data
Data from Johns Hopkins database on Covid data from 2020-2023
```{r get_data}
## Get current data in the four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)

global_deaths <- read_csv(urls[4])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
us_cases <- read_csv(urls[1])
```

## Tidy Data
```{r tidy_data}
us_deaths <- us_deaths %>%
     pivot_longer(cols = 
          -c(`Province_State`, `Country_Region`, Lat, Long_,
             UID, iso2, iso3, code3, FIPS, Admin2, Combined_Key),
            names_to = "date",
            values_to = "deaths") %>%
     select(-c(Lat, Long_))

global_cases <- global_cases %>%
     pivot_longer(cols = 
          -c(`Province/State`, `Country/Region`, Lat, Long),
            names_to = "date",
            values_to = "cases") %>%
     select(-c(Lat, Long))

us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  select(-c(Lat, Long_))
us_cases <- us_cases %>% filter(date != "Population" & cases > -1)

global_deaths <- global_deaths %>%
  pivot_longer(cols = -(`Province/State`:Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global_deaths$date <- as.Date(global_deaths$date, format = "%m/%d/%y")

```
Data cleaned to simplify one row per date per location (dependent on table)
Cleaning for tables as needed to adjust dates into correct formats, clean out unnecessary columns, etc
Tables provide a cumulative case/death count grouped by location over time

## Visualize Data
```{r graph_data}
summed_us_deaths <- us_deaths %>% rename(state = Province_State) %>% filter(date == "3/9/23") %>% group_by(state) %>% summarise(total_deaths = sum(deaths, na.rm = TRUE))
plot_usmap(data = summed_us_deaths, values = "total_deaths", color = "red") +
  scale_fill_continuous(name = "Covid Deaths") + 
  theme(legend.position = "right") +
  labs(title = "Total Covid Deaths by State")

top_countries <- c("US", "Russia", "India", "Indonesia", "Pakistan", "China")
filtered_global <- global_deaths %>%
  filter(`Country/Region` %in% top_countries)
ggplot(filtered_global, aes(x = date, y = deaths, color = `Country/Region`, group = `Country/Region`)) + 
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = function(y) format(y, scientific = FALSE)) +
  labs(title = "Cumulative Covid Deaths by top Countries")
```
The first graph (the map of the US) is exactly what you'd expect to find: the higher population states have more cases. Texas, California, New York and Florida, the highest population states, also have the most cases reported.

The second graph has some more interesting conclusions to draw. The US blows the rest of the world out of the water in terms of Covid deaths. I filtered the data here for some of the most populous countries and also some of the countries with highest death counts, and there were some anomalies. China is one of the top populated countries on earth (and where Covid originated), but it has remarkably low death counts. Interesting. Same story for Pakistan. Russia is a far less populous country, though it has double the cases of Indonesia despite its lower population and far lower population density. This is likely due to differences in reporting more than anything, to be touched on in the bias discussion.

## Model Data
```{r model_data}
pop_table <- us_deaths %>%
  filter(date %in% c("Population", "3/9/23")) %>%
  pivot_wider(names_from = date, values_from = deaths) %>%
  rename(deaths = `3/9/23`)
model <- lm(deaths ~ Population, data = pop_table)
summary(model)
```
Here I wanted to see just how much of an impact population had on death counts. The US data was set up to be more reliable than the world data so I ran with it. Turns out, with an R^2 value of .89 (89% of variance due to population) and a p value of 2e^-16, there is a ridiculously strong correlation between the two. I was expecting it to be relatively strong but those are astronomical numbers. Population almost solely dictates how many cases a county would get.

## Conclusion/Bias
In conclusion, population of a country/state has a major impact on the number of cases or deaths said region would have. That makes sense. The more people you have, the higher potential for total cases or deaths you have as well.
That said, this is complicated by reporting bias. A country is not necessarily incentivized to fully admit to total death/case numbers. There is no international police force to crack down on statistical reporting. That leads to inflated (well, deflated) case counts and death tolls reported in countries you'd expect to see far higher numbers from. China hovering in just the 10s of thousands of deaths while the US had well over a million suggests something is off. Especially considering the high correlation bewteen population and deaths within the states (which had more regular and reliable reporting numbers).